A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-26 article/02_Materials_and_Methods.Rmd
language in materials&methods

To create a tidy and useful subset of Radon-B, the following steps of data preparation were applied. Please see the code for details of implementation. The value in brackets indicates the number of dates still available after the respective filtering operation.

@@

7. Manually defining a rectangular (EPSG:102013) research area with sufficient information in Central-, Northern and Northwestern Europe based on perceived spatial data distribution (Figure \ref{map_graves}).

@@

10. Transforming data from the meaning *^14^C dates on graves* to *burial type at a certain time*. For `r r("bronze15 multi dates one grave", sf_prep)` entries in the thus far assembled subset, the attribution to *site* and *feature* is not unique. That means that for many graves multiple (up to 8) ^14^C dates have been created and entered in Radon-B. Dates cannot be treated as semantically equal to graves. There's a variety of phenomena that cause this effect: In some cases, safety measurements were taken -- multiple dates for one single grave that all fall within the margin of errors of each other. In other cases, there is a set of strongly diverging dates for a collective grave that was used for hundreds of years. There are also cases where the *feature* information in the database is not sufficient to determine whether two dates are really from the same grave despite the identical entry. The following steps were applied to solve the problem automatically. This approach is compatible with the intention to create a cultural trait time series as explained below. (`r r("size bronze17", sf_prep)` graves)

@@

11. Defining macro-regions to observe the development of *burial type* and *burial construction* within sensible geographical and cultural units that are appropriate to the amount of available data. The regions are circular and were delimited semi-automatically by creating a regular grid in accordance to main cultural-historical regions of the European Bronze Age [@fokkensIntroductionBronzeAge2013]. The regions have a radius of 240km and overlap slightly (Figure \ref{map_regions} A). That means some graves are counted towards not just one but two regions.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../analysis/figures/map_regions.jpeg}
\caption{\label{map_regions}Circular, artificial regions (\textbf{A}) and spatial distance network with classified distance values (\textbf{B}). The map layout is the same as in Figure \ref{map_graves}.}
\end{figure}

The resulting dataset is structured as a table with `r r("gpr size", sf_desc)` rows each representing one grave as compiled by `r r("dpr size", sf_desc)` ^14^C dates from `r r("dpr sites amount", sf_desc)` sites. The graves are linked to `r r("dpr period amount", sf_desc)` different *period* terms and `r r("dpr culture amount", sf_desc)` different archaeological *cultures* as they are distinguished in Radon-B. `r r("dpr material bone amount", sf_desc)` dates were measured on sample material of bones and teeth (at least `r r("dpr material cremated bones amount", sf_desc)` of them burnt), `r r("dpr material charcoal wood amount", sf_desc)` on wood and charcoal, a small part (`r r("dpr material other amount", sf_desc)`) of other materials such as nuts, resin or pitch. For the remaining `r r("dpr material unknown amount", sf_desc)` dates, no material information is stored. Regarding the variables burial type and burial construction, the conditions shown in Table \ref{tab:dprcrosstab} apply.

>>>>

To create a tidy and useful subset of Radon-B, the following steps of data preparation were applied. Please see the code for details of implementation. The value in brackets indicates the number of dates or graves still available after the respective filtering operation.

@@

7. Manually defining a rectangular (in EPSG:102013) research area with sufficient information in Central-, Northern and Northwestern Europe based on perceived spatial data distribution (Figure \ref{map_graves}).

@@

10. Transforming data from the meaning *^14^C dates on graves* to *burial type at a certain time*. For `r r("bronze15 multi dates one grave", sf_prep)` entries in the thus far assembled subset the attribution to *site* and *feature* is not unique. That means that for many graves multiple (up to 8) ^14^C dates have been created (and entered in Radon-B) and dates therefore cannot be treated as semantically equal to graves. There is a variety of phenomena that cause this effect: In some cases, safety measurements were taken so that there are multiple dates for one single grave that all fall within the margin of errors of each other. In other cases, the set of dates for one grave strongly diverge because they cover the life cycle of a collective grave that was used for hundreds of years. There are also cases where the *feature* information in Radon-B is not sufficient to determine whether two dates are really from the same grave despite the identical entry. The following steps were applied to solve this problem algorithmically. The approach is compatible with the intention to create a cultural trait time series as explained below. (`r r("size bronze17", sf_prep)` graves)

@@

11. Defining macro-regions to observe the development of *burial type* and *burial construction* within sensible geographical and cultural units that are appropriate to the amount of available data. The regions are circular and were delimited semi-automatically by creating a regular grid in accordance to main cultural-historical regions of the European Bronze Age [@fokkensIntroductionBronzeAge2013]. The regions have a radius of 240km and overlap slightly (Figure \ref{map_regions} A). That means some graves are counted towards not only one but two regions.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../analysis/figures/map_regions.jpeg}
\caption{\label{map_regions}Circular, artificial regions (\textbf{A}) and spatial distance network with classified distance values (\textbf{B}). The map layout is similar to Figure \ref{map_graves} but instead of rivers modern country borders are drawn.}
\end{figure}

The resulting dataset is structured as a table with `r r("gpr size", sf_desc)` rows each representing one grave as compiled by `r r("dpr size", sf_desc)` ^14^C dates from `r r("dpr sites amount", sf_desc)` sites. The graves are linked to `r r("dpr period amount", sf_desc)` different *period* terms and `r r("dpr culture amount", sf_desc)` different archaeological *cultures* as they are distinguished in Radon-B. `r r("dpr material bone amount", sf_desc)` dates were measured on sample material of bones and teeth (at least `r r("dpr material cremated bones amount", sf_desc)` of them burnt), `r r("dpr material charcoal wood amount", sf_desc)` on wood and charcoal, a small part (`r r("dpr material other amount", sf_desc)`) on other materials such as nuts, resin or pitch. For the remaining `r r("dpr material unknown amount", sf_desc)` dates, no material information is stored. Regarding the variables burial type and burial construction, the conditions shown in Table \ref{tab:dprcrosstab} apply.
